# BatWing
___
```
Package write in the R programming language
```
```
Version 1.0.0. Beta (2016-06-10)
```


This package calculates the Lifting Surface Area (LSA) of a Bat's wing using different formulas, and also calculate the dynamics measures of the Bat's wing.

The input files for both LSA function and Dynamics function are very specific, the colums order must be the same as the example data ( https://github.com/oleon12/BatWing/tree/master/data ).

___

## Installation

```{r}
library("devtools")

install_github("oleon12/BatWing")
```
___

## Formulas

The LSA formulas which this package use are the proposal by:

- **Pirlot 1977**
  
  ![P](https://latex.codecogs.com/gif.latex?LSA%20%3D%200.735%20*B%20*V)
- **Smith & Starrett 1979**
  
  ![SS](https://latex.codecogs.com/gif.latex?LSA%20%3D%202*%5B%28FA*V%29&plus;%5Cfrac%7B1%7D%7B2%7D%20*%28cos%28%5Calpha%20%29*III%20*IV%29&plus;%5Cfrac%7B1%7D%7B2%7D*%28sin%28%5Calpha%29*IV%20*V%29%5D)

  where:

  ![A](https://latex.codecogs.com/gif.latex?%5Calpha%20%3D%20arctan%28%5Cfrac%7BV%7D%7BIII%7D%29)

- **Blood & McFarlane 1988**

  ![BM](https://latex.codecogs.com/gif.latex?LSA%20%3D%202%20*%20%5B%28FA*V%29&plus;%5Cfrac%7B1%7D%7B2%7D*%28V*III%29%5D)

- **Aldrige 1988**

  ![AA](https://latex.codecogs.com/gif.latex?LSA%20%3D%20%282.85*10%5E%7B-3%7D%29*M%5E%7B%5Cfrac%7B2%7D%7B3%7D%7D)

   ___
 
   <sup>*Where:*

   <sup> _**FA** = Forearm Length_, _**B**= Wingspan_ 
   , _**III** = Digit 3 Length_, _**IV** = Digit 4 Length_
   , _**V** = Digit 5 Length_, _**M** = Mass*_
  ___ 
   
The Dynamics formulas which this package use are:

- **Aspect Ratio (Vaughan 1959; Norberg & Rayner 1987)**

  ![AR](https://latex.codecogs.com/gif.latex?AR%20%3D%20%5Cfrac%7BB%5E%7B2%7D%7D%7BS%7D)
  
- **Wing Loading (Norberg & Rayner 1987)**
  
  ![WL](https://latex.codecogs.com/gif.latex?WL%20%3D%20%5Cfrac%7B%28M*9.81%5Cfrac%7Bm%7D%7Bsec%5E%7B2%7D%7D%29%7D%7BS%7D)
  
- **Relative Wing Loading (Norberg 2000)**
  
  ![RWL](https://latex.codecogs.com/gif.latex?RWL%3D%5Cfrac%7BWL%7D%7BM%5E%7B%5Cfrac%7B1%7D%7B3%7D%7D%7D)
  
- **Minimum Power Speed (Norber & Rayner 1987)**
  
  ![MPS](https://latex.codecogs.com/gif.latex?MPS%3D6.58%20*M%5E%7B0.422%7D*B%5E%7B-0.479%7D*S%5E%7B-0.148%7D)
  
  ___
 
   <sup>*Where:*

   <sup> _**S** = LSA in squared meters_, _**B**= Wingspan in squared meters_ 
   ,  _**9.81** = Gravity constant_, _**M** = Mass in Kilograms*_
  ___ 
  
___

## Example
___

### BatWingArea()

BatWingArea function calculate the Lifting Surface Area of a bat wing using different wing measures (see Formulas section).
This function also calculate the mean, standard deviation and variance of the input variables and LSA measures.

Load the library and the example data **"wing"**, then check what is in the data example.

```{r}
library(BatWing)

data(wing)
head(wing)
```
The data contains the identification colum (id)(in this case colector acronym, but also could be species name or other name),
Forearm length, length of the digits 5, 4 and 3, mass and wingspan. All measures are in milimeters and grames. The id colum always must be a colum, never row names. The variables sort of the input data must be the same as this example data, if not, the calculation will be wrong.

```
         id ForearmLength Digit5.Length Digit4.Length Digit3.Length Mass Wingspan
1 ODLAM-011          41.4         63.20         61.70         87.50 15.5      330
2 ODLAM-010          42.4         64.10         62.60         94.90 17.5      325
3 ODLAM-012          37.9         54.60         58.20         76.00  9.0      295
4 ODLAM-014          41.5         58.30         58.95         79.70 16.5      319
5 ODLAM-016          31.0         28.40         49.90         60.75  7.5      238
6 ODLAM-018          42.1         59.15         61.90         79.90 15.5      331
```

Now, calculate the LSA for the wing data. Scale argument transform the variables from milimeters and grames to meters and kilograms, so you can have the LSA in squared meters or squared milimeters

The outcome **_LSA_** is a S3 object with two data frames : 1. Area (the LSA measures) and 2. Stats (variables statistics).
Let's take a look to the LSA measures.

```{r}
LSA <- BatWingArea(x=wing, scale=F)
head(LSA$Area)
```

The function calculated four different LSA formulas: BM: Blood & McFarlane 1988, A: Aldrige 1988, SS: Smith & Starrett 1979 and P: Pirlot 1977 (detailed formulas in Formulas section). 

```
                 BM         A        SS         P
ODLAM-011 10762.960 0.2282375 11892.700 15329.160
ODLAM-010 11518.770 0.2909375 12604.629 15311.887
ODLAM-012  8288.280 0.0769500  9585.018 11838.645
ODLAM-014  9485.410 0.2586375 10660.039 13669.309
ODLAM-016  3486.100 0.0534375  5107.123  4968.012
ODLAM-018  9706.515 0.2282375 11134.029 14390.308
```

Now, let's take a look to the variables statistics.

```{r}
LSA$Stats
```

```
     ForearmLength Digit5.Length Digit4.Length Digit3.Length      Mass   Wingspan           BM         A           SS            P
mean      43.60000      55.67000      62.89500      84.46500 17.750000  337.30000    10021.071 0.3657263    11733.002    14470.143
sd        10.87288      16.09049      13.36127      17.12325  8.813657   77.11326     4945.237 0.4191581     5577.896     7309.019
var      118.21944     258.90400     178.52358     293.20558 77.680556 5946.45556 24455364.257 0.1756935 31112928.033 53421752.534
```

```{r}
LSA <- BatWingArea(x=wing, scale=F, method=c("BM","SS"))
head(LSA$Area)
```

```
                 BM        SS
ODLAM-011 10762.960 11892.700
ODLAM-010 11518.770 12604.629
ODLAM-012  8288.280  9585.018
ODLAM-014  9485.410 10660.039
ODLAM-016  3486.100  5107.123
ODLAM-018  9706.515 11134.029
```
___
### BatWingDynamics()

```{r}
data(dynamics)
head(dynamics)
```

```
         ID Mass Wingspan Area(inM^2)
1 ODLAM-011 15.5      330 0.011892700
2 ODLAM-010 17.5      325 0.012604629
3 ODLAM-012  9.0      295 0.009585018
4 ODLAM-014 16.5      319 0.010660039
5 ODLAM-016  7.5      238 0.005107123
6 ODLAM-018 15.5      331 0.011134029
```

```{r}
Dym <- BatWingDynamics(x=dynamics, scale=T)
head(Dym$Dynamics)
```

```
                 AR       WL      RWL      MPS
ODLAM-011  9.156878 12.78557 51.27941 3.715662
ODLAM-010  8.379858 13.62000 52.46032 3.905866
ODLAM-012  9.079273  9.21125 44.28307 3.218072
ODLAM-014  9.546025 15.18428 59.64392 3.940761
ODLAM-016 11.091175 14.40635 73.59814 3.625027
ODLAM-018  9.840194 13.65678 54.77358 3.746655
```

```{r}
Dym$Stats
```

```
             Mass    Wingspan  Area(inM^2)        AR       WL       RWL       MPS
mean 1.775000e-02 0.337300000 1.173300e-02 10.333714 15.48375  61.14137 3.8564452
sd   8.813657e-03 0.077113264 5.577896e-03  1.752498  5.16755  20.17576 0.3835981
var  7.768056e-05 0.005946456 3.111293e-05  3.071250 26.70357 407.06138 0.1471475
```

```{r}
Dym <- BatWingDynamics(x= dynamics, scale=T, method = c("AR","RWL"))
head(Dym$Dynamics)
```

```
                 AR      RWL
ODLAM-011  9.156878 51.27941
ODLAM-010  8.379858 52.46032
ODLAM-012  9.079273 44.28307
ODLAM-014  9.546025 59.64392
ODLAM-016 11.091175 73.59814
ODLAM-018  9.840194 54.77358
```
___
### BatWingAll()

```{r}
data(wing)

All <- BatWingAll(x=wing, scale=T)

All$Dynamics
```
```
$BloodMcFarlane
                 Area        AR       WL       RWL      MPS
ODLAM-011 0.010762960 10.118034 14.12762  56.66198 3.770959
ODLAM-010 0.011518770  9.169816 14.90394  57.40569 3.958291
ODLAM-012 0.008288280 10.499766 10.65239  51.21135 3.288053
ODLAM-014 0.009485410 10.728160 17.06463  67.02994 4.009443
ODLAM-016 0.003486100 16.248530 21.10525 107.82100 3.835792
ODLAM-018 0.009706515 11.287367 15.66525  62.82900 3.823516
ODLAM-019 0.007846080 11.778621 23.75581  89.02637 4.478730
ODLAM-020 0.021146895 12.347959 18.55592  54.25783 4.128812
ODLAM-021 0.013487540 13.518329 14.91043  54.48022 3.627280
ODLAm-022 0.004482160 19.153488 36.11317 141.85270 4.666104

$Aldrige
                  Area        AR        WL       RWL      MPS
ODLAM-011 2.282375e-07  477134.6  666213.9 2672000.1 18.54114
ODLAM-010 2.909375e-07  363050.5  590075.2 2272800.6 18.96503
ODLAM-012 7.695000e-08 1130929.2 1147368.4 5515972.0 18.26891
ODLAM-014 2.586375e-07  393450.3  625837.3 2458292.0 18.99372
ODLAM-016 5.343750e-08 1060004.7 1376842.1 7033913.9 19.78792
ODLAM-018 2.282375e-07  480030.7  666213.9 2672000.1 18.51429
ODLAM-019 3.429500e-07  269473.7  543490.3 2036763.5 19.78560
ODLAM-020 1.520000e-06  171790.1  258157.9  754858.3 16.94528
ODLAM-021 3.992375e-07  456693.1  503722.7 1840518.8 16.97574
ODLAm-022 2.586375e-07  331927.9  625837.3 2458292.0 19.78318

$SmithStarrett
                 Area        AR       WL       RWL      MPS
ODLAM-011 0.011892700  9.156878 12.78557  51.27941 3.715662
ODLAM-010 0.012604629  8.379858 13.62000  52.46032 3.905866
ODLAM-012 0.009585018  9.079273  9.21125  44.28307 3.218072
ODLAM-014 0.010660039  9.546025 15.18428  59.64392 3.940761
ODLAM-016 0.005107123 11.091175 14.40635  73.59814 3.625027
ODLAM-018 0.011134029  9.840194 13.65678  54.77358 3.746655
ODLAM-019 0.009181822 10.065105 20.29989  76.07510 4.375725
ODLAM-020 0.024440866 10.683787 16.05508  46.94533 4.041293
ODLAM-021 0.016881464 10.800544 11.91277  43.52727 3.508766
ODLAm-022 0.005842332 14.694303 27.70555 108.82752 4.486625

$Pirlot
                 Area        AR        WL      RWL      MPS
ODLAM-011 0.015329160  7.104107  9.919330 39.78369 3.578664
ODLAM-010 0.015311887  6.898235 11.211877 43.18494 3.794998
ODLAM-012 0.011838645  7.350926  7.457779 35.85326 3.119053
ODLAM-014 0.013669310  7.444487 11.841491 46.51343 3.798375
ODLAM-016 0.004968012 11.401744 14.809747 75.65899 3.639874
ODLAM-018 0.014390308  7.613527 10.566487 42.37926 3.607064
ODLAM-019 0.011797632  7.833436 15.798933 59.20748 4.216361
ODLAM-020 0.031605478  8.261891 12.415569 36.30334 3.890423
ODLAM-021 0.019050391  9.570879 10.556476 38.57160 3.446556
ODLAm-022 0.006740612 12.736085 24.013400 94.32475 4.392654
```

```{r}
All <- BatWingAll(x=wing, scale=F)
```

```
Warning message:
In BatWingAll(bat.m, scale = F) :
  scaleArea is FALSE, remember, for the Dynamic formulas the LSA or Area must be in squared meters
```
